The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2022-01-23 paper/paper.tex
updated compiled paper

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\begin{document}
\maketitle

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021).

This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.

This disparity between the average and the individual is a part of statistics that is given less attention. It is this disparity which makes this particular data set very important as a textbook example. Textbook data sets have staying power if it has a challenge that has never been completely solved. The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?

@@

As a textbook data though, the dates are tired. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets sets up a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This is our story of the journey from open (Open Knowledge Foundation 2021) but wild data to textbook data, it's not entirely pretty.

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021. {``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH.

@@

>>>>

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\def\tightlist{}

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021).

This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.

This disparity between the average and the individual is a part of statistics that is given less attention. It is this disparity which makes this particular data set important as a textbook example. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?

@@

As a textbook data though, the original data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets sets up a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/etan0038/Dropbox/repo/yowie-paper/paper/figures/flowchart_blind}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.

\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021. {``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH.

@@

\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
